### 2015-03-16 doc/thesis/typedlua.tex
Minor improvements to the first part of Chapter 3

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them on the type system.
However, this evaluation is out of the scope of this work.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
As we mentioned in the previous chapter,
a sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
These examples give an informal overview of our optional type system.
We will use typing rules to present the formalization of our optional type
system in the next chapter.
All the examples that we present in this chapter run in our Typed Lua compiler.

@@

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
Union types are supertypes of their parts.
The base types \texttt{boolean}, \texttt{integer}, \texttt{number},
and \texttt{string} are supertypes of their respective literal types.
The base type \texttt{integer} is subtype of \texttt{number}.
Function types are related by contravariance on the input
and covariance on the output.
Table types have width subtyping, with depth subtyping on
\texttt{const} fields.
Tuple and vararg types are covariant.
Unions of tuple types are also supertypes of their parts.

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
neither a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and we can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.

@@

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
In fact, we can overload the minus operator to return a value that is
not even a number.
However, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code break the
guarantees provided by the statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

In the default mode, the type of \texttt{sum} is
\texttt{(integer, integer, value*) -> (integer, nil*)}.
The compiler adds \texttt{value*} to the type of the parameter list
to discard extra arguments.
For instance, the call \texttt{sum(1, 2, 3)} compiles without any
warnings because Typed Lua uses \texttt{value*} to drop the extra
argument \texttt{3}.

In the strict mode, the type of \texttt{sum} is
\texttt{(integer, integer) -> (integer)}.
The compiler does not add \texttt{value*} to the type of the parameter
list to catch arity mismatch.
For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
because we are passing an extra argument \texttt{3} to \texttt{sum}.

@@

@@

Typed Lua allows the assignment \texttt{v = v or e} to refine
the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{v} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{v} until the
next assignment to \texttt{v}.
In the case of \texttt{greeting}, the assignment constrains its type
from \texttt{string|nil} to \texttt{string}.

@@

The type of \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The call \texttt{message("Typed Lua")} compiles without
any warnings in both modes, because the argument list
\texttt{("Typed Lua", nil*)} is consistent with
\texttt{(string, string|nil, value*)} in the default mode,
and it is also consistent with  
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(v) == "string"} predicate that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} inside the \texttt{if} body and to \texttt{t}
inside the \texttt{else} body.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and returns \texttt{nil} plus an error message or other data that
describes the error when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) // divisor
        return q, r
      end 
    end
\end{verbatim}

@@

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.
Later in this chapter we will show that we use the syntax of records
to type modules and objects.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
unique, otherwise it is closed.
Any alias to unique variables makes them open, as we use the
tag open to keep track of unique variables that are aliased.
In the previous example, \texttt{person} has an unique type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
fixed to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

Typed Lua uses three rules to decide whether it is sound to
allow field assignment to change the type of unique and open
variables:
the variable must be local to the current block,
the new type must only add new fields, and
the variable cannot have been assigned in another function.

@@

We can create aliases to open variables, but they are always closed.
Any mutation in the original reference and in the aliased reference
is not a problem, as the type of the original reference can only
add new fields.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.
In our previous example, the second line changes the type of
\texttt{person} from unique to open and makes the type of
\texttt{bogus} closed.

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string,
                              "lastname":string}
\end{verbatim}

Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
for table types that specify records, as writing table types can be unwieldy
when records get bigger and the types of record fields get more complicated.
Figure \ref{fig:interfaces} shows the concrete syntax of the
\texttt{interface} construct.
In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
declare the type of classes, and that \emph{methodtype} is syntactic
sugar to express the type of methods.

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring table types that express records.
For this reason, it worth having two different constructs;
one specifically for records and another for more general types.

@@

An idiomatic way for defining modules in Lua is to declare only locals
and return a table constructor at the end of the source file.
The returned constructor includes the members that the module should export.
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

In this example, Typed Lua uses the type of the table constructor
to allow the programmer to build the type of the module.
The local variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we did not include it in the list of exported members. 
The \texttt{return} at the end of the source file gives the type of the module:
\begin{verbatim}
    { "deg":(number) -> (number),
      "rad":(number) -> (number),
      "pow":(number, number) -> (number) }
\end{verbatim}

Another idiomatic way for defining modules in Lua is to declare an
empty table at the begin of the source file,
populate this table with the members that the module should export,
and return this table at the end of the source file. 
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local mymath = {}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    mymath.deg = function (x:number)
      return x / RADIANS_PER_DEGREE
    end
    mymath.rad = function (x:number)
      return x * RADIANS_PER_DEGREE
    end
    mymath.pow = function (x:number, y:number)
      return x ^ y
    end
    return mymath
\end{verbatim}

In this example, Typed Lua uses the refinement of table types to
allow the programmer to incrementally build the type of the module.
The variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we declared it as local to the module. 
The \texttt{return} at the end of the source file gives the
type of the module, which is the same type of \texttt{mymath}.
This module has the same type of the module from the previous example.

@@

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer the type of this module.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements the given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

The way that Typed Lua handles modules using table types is also
relevant for supporting object-oriented programming, as we discuss in
the next section.

@@

>>>>

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them in the type system.
However, this evaluation is out of scope of this work.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
A sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
These examples give an informal overview of our optional type system.
In the next chapter we will use typing rules to present the formalization
of the most interesting features of our optional type system.
All the examples that we present in this chapter run in our Typed Lua compiler.

@@

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
Union types are supertypes of their parts.
The base types \texttt{boolean}, \texttt{integer}, \texttt{number},
and \texttt{string} are supertypes of their respective literal types.
The base type \texttt{integer} is subtype of \texttt{number}.
Function types are related by contravariance on the input
and covariance on the output.
Table types have width subtyping, with depth subtyping on
\texttt{const} fields.
Tuple and vararg types are covariant.
Unions of tuple types are also supertypes of their parts.
We will present the formalization of the subtyping relation
in Section \ref{sec:subtyping}.

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
neither a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and we can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.
We will discuss the formalization of the consistent-subtyping
relation in Section \ref{sec:subtyping}.

@@

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
In fact, we can overload the minus operator to return a value that is
not even a number.
However, we can call \texttt{absolute} inside \texttt{distance}
because the subtraction expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code break the
guarantees provided by the statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

In the default mode, \texttt{sum} has type
\texttt{(integer, integer, value*) -> (integer, nil*)}.
The compiler adds \texttt{value*} to the type of the parameter list
to discard extra arguments.
For instance, the call \texttt{sum(1, 2, 3)} compiles without any
warnings because Typed Lua uses \texttt{value*} to drop the extra
argument \texttt{3}.

In the strict mode, \texttt{sum} has type
\texttt{(integer, integer) -> (integer)}.
The compiler does not add \texttt{value*} to the type of the parameter
list to catch arity mismatch.
For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
because we are passing an extra argument \texttt{3} to \texttt{sum}.

@@

As another example,
\begin{verbatim}
    local x:integer, y:integer = sum(2, 2)
\end{verbatim}
compiles with warnings in both operation modes.
This means that the left-hand side of this local declaration has
the tuple type \texttt{(integer, integer, value*)}, while the
expression list in the right-hand side of this local declaration
has the tuple type \texttt{(integer, nil*)} in both operation modes.
The type \texttt{nil*} generates a value of type \texttt{nil}
that is not consistent with \texttt{integer}, the type of \texttt{y}.

@@

Typed Lua allows the assignment \texttt{v = v or e} to change
the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{v} is local to the current function
and it is not being assigned in another function.
The change only affects the type of \texttt{v} in the remainder
of the current scope.
Any assignment to \texttt{v} restores its type back to the
original type.
In the case of \texttt{greeting}, the assignment change its type
from \texttt{string|nil} to \texttt{string}.

@@

The type of \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The call \texttt{message("Typed Lua")} compiles without
any warnings in both modes, because the argument list
\texttt{("Typed Lua", nil*)} is consistent with the input type
\texttt{(string, string|nil, value*)} in the default mode,
and it is also consistent with the input type 
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.
In Section \ref{sec:assignment} we will formalize this behavior.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the predicate \texttt{type(v) == "string"} that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and it returns \texttt{nil} plus an error message or other data that
describes the error when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) // divisor
        return q, r
      end 
    end
\end{verbatim}

@@

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.

Even though Typed Lua includes two different forms for annotating
a table, in this section we only discuss the form
$\texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}$.
We call this form \emph{closed} table type because it
represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In Sections \ref{sec:modules} and \ref{sec:oop} we will show that
we use the syntax $\texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}$
to type modules and objects.
We call this form \emph{fixed} table type because it
represents tables with no keys that do not inhabit one of the table's
key types, and can have safe aliases.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.
In the next chapter we formalize and discuss in more detail the
different table types that our type system has.

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
\emph{unique}, otherwise it can be either \emph{fixed} or
\emph{closed}.
In this section we discuss the cases that produce \emph{closed}
table types, while in Sections \ref{sec:modules} and \ref{sec:oop}
we will discuss the cases that produce \emph{fixed} table types
to describe the type of modules and classes.
Any alias to \emph{unique} variables makes them \emph{open}, as we use the
tag \emph{open} to keep track of \emph{unique} variables that are aliased.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

Typed Lua uses three rules to decide whether it is sound to
allow field assignment to change the type of \emph{unique} and
\emph{open} variables:
the variable must be local to the current block,
the new type must only add new fields, and
the variable cannot have been assigned in another function.

@@

We can create aliases to \emph{open} variables, but they are always
\emph{closed}.
Any mutation in the original reference and in the aliased reference
is not a problem, as the type of the original reference can only
add new fields.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.
In our previous example, the second line changes the type of
\texttt{person} from \emph{unique} to \emph{open} and makes the type of
\texttt{bogus} \emph{closed}.

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the \emph{closed} table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string,
                              "lastname":string}
\end{verbatim}

Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
for \emph{closed} table types that specify records, as writing table types can be unwieldy
when records get bigger and the types of record fields get more complicated.
Figure \ref{fig:interfaces} shows the concrete syntax of the
\texttt{interface} construct.

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{type}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring \emph{closed} table types that express records.
For this reason, it worth having two different constructs;
one specifically for records and another for more general types.

@@

An idiomatic way for defining modules in Lua is to declare only locals
and return a table constructor at the end of the source file.
The returned constructor includes the members that the module should export.
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number):number
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number):number
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number):number
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

In this example, Typed Lua uses the type of the table constructor
to allow the programmer to build the type of the module.
The local variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we did not include it in the list of exported members. 
The \texttt{return} at the end of the source file gives the type of the module:
\begin{verbatim}
    <{ "deg":(number) -> (number),
       "rad":(number) -> (number),
       "pow":(number, number) -> (number) }>
\end{verbatim}

Typed Lua always fixes \emph{unique} and \emph{open} table types
that appear in a return statement.
Typed Lua has this behavior to ensure that modules and classes
always implement \emph{fixed} table types.
In the next section we will show that we build classes in a
similar way that we use to build modules, and we will also show
that classes should have \emph{fixed} table types to allow safe
inheritance.

Another idiomatic way for defining modules in Lua is to declare an
empty table at the begin of the source file,
populate this table with the members that the module should export,
and return this table at the end of the source file. 
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local mymath = {}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    mymath.deg = function (x:number):number
      return x / RADIANS_PER_DEGREE
    end
    mymath.rad = function (x:number):number
      return x * RADIANS_PER_DEGREE
    end
    mymath.pow = function (x:number, y:number):number
      return x ^ y
    end
    return mymath
\end{verbatim}

In this example, Typed Lua uses the refinement of table types to
allow the programmer to incrementally build the type of the module
\texttt{mymath}.
The variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we declared it as local to the module. 
The \texttt{return} at the end of the source file gives the
type of the module, which is the same type of the variable \texttt{mymath}.
This module has the same type of the module from the previous example.

@@

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer the type of this module.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements a given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

The way that Typed Lua handles modules using \emph{fixed} table types
is also relevant for supporting object-oriented programming,
as we discuss in the next section.

@@

\begin{figure}[!ht]
\textbf{Classes}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:classes}
\end{figure}
